\section{Bid Info}

Application:  https://ccfrms.nihr.ac.uk/

 
\begin{markdown}

* Call opens – 17 May 2023    \item Expression of Interest - (500 words) 16 August 2023 at 1pm
* Close call – 13 September at 1pm
* Stage 1 outcome – December 2023
* Stage 2 call close - end of January 2024
* Stage 2 outcome – May 2024
* Start - September 2024?

## What does RfPB cover?

* Research into the provision and use of NHS services.
* Effectiveness and cost effectiveness evaluations of interventions.
* Research that examines the resource use of alternative means for healthcare delivery.
* Feasibility research to support applications for major awards to other funders.
* Development and refining of new interventions, scales or outcome measures.
*  Research to explore the potential for improving patient health and wellbeing through needs assessments, methods development and exploratory studies.
* Evidence synthesis and systematic reviews


Links:

Main page: https://www.nihr.ac.uk/funding/rfpb-under-represented-disciplines-and-specialisms-highlight-notice-methodologists/33231

Spec: https://www.nihr.ac.uk/documents/rfpb-under-represented-disciplines-and-specialisms-highlight-notice-methodologists-call-specification/33366

FAQ: https://www.nihr.ac.uk/documents/rfpb-under-represented-disciplines-and-specialisms-highlight-notice-methodologists-frequently-asked-questions/33363

Expression of interest: https://docs.google.com/forms/d/e/1FAIpQLSdmNhKi-k5HN9MeRRyKxoxTAqt-wIv8JbOz4kttXSHyEczKfQ/viewform

Application portal: https://ccfrms.nihr.ac.uk/Login.aspx?ReturnUrl=%2f

Stage 1 Guidance: https://www.nihr.ac.uk/documents/research-for-patient-benefit-programme-guidance-for-stage-1-applications/20154

Stage 2 Guidance: https://www.nihr.ac.uk/documents/research-for-patient-benefit-guidance-for-stage-2-applications/20556

Stage 1 template: https://www.nihr.ac.uk/documents/funding/cross-programme/NIHR-Template-Application-Form-stage-1.docx

Stage 2 template: https://www.nihr.ac.uk/documents/funding/cross-programme/NIHR-Template-Application-Form-stage-2.docx


\end{markdown}

Project ID: 2692263
